A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2017-01-07 imaging_demo/ms.tex
Minor updates throughout
From b5e83c5 (mnras):
====================

In ms.tex,

(*) Specify data used is from LWA1
(*) Minor changes throughout

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. \sout{It takes advantage of the multiplication-convolution theorem of Fourier transforms.} Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. EPIC is highly modular, parallelizable, implemented in object-oriented Python, and publicly available. We have verified the images produced to be equivalent to those from traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA). {\bf We provide a detailed framework for imaging with heterogeneous arrays and show that EPIC} robustly estimates the input sky model for {\bf such} arrays. Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying an optimized version of EPIC. The algorithm is a strong candidate for instruments targeting transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena due to the availability of high-speed calibrated time-domain images and low output bandwidth relative to visibility-based systems.
\end{abstract}

@@

This paradigm shift requires a fundamentally new approach to the design of digital correlators \citep{lon00}. Modern correlators calculate the cross-power correlation between all antenna pairs in many narrow frequencies, forming \emph{visibilities}, the fundamental measurement of traditional radio interferometers. The computational requirements for a modern FX correlator scale with the number of antenna pairs, or the square of the number of antennas $\sim \Nant^2$ \citep{bun04}. For this reason traditional correlators have difficulty scaling to thousands of antennas. As an example, the full HERA correlator for 352 dishes with 200 MHz of bandwidth requires 212 trillion complex multiply-accumulates per second (TMACS). Future arrays with thousands of collecting elements will require orders of magnitude more computation, making the correlator the dominant cost.

@@

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA. {\bf In \S\ref{sec:versatility}, we present a mathematical framework for imaging with heterogeneous arrays and demonstrate the ability of EPIC to robustly image data from such arrays}. In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator is computationally more efficient, and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{i 2\pi f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \nonumber\\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-i 2\pi f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. The noise-like cross-terms ideally have zero mean. This expression can be further simplified using the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)$, as $\left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t = \mathcal{I}(\hat{\mathbf{s}},f,t)\,\delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{B}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power pattern, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to the measurement plane, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

@@

Hereafter, we adopt the matrix notation of \citet{mor11}, where vectors are represented with single coordinates, and matrices are represented by two coordinates denoting the spaces the operator transforms between. Since each frequency is processed independently, we drop the dependence on $f$. In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \mathbf{B}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}})$ is Fourier transformed using $\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence function is weighted and summed using the antenna power response, $\mathbf{B}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \mathbf{E}(\mathbf{a})\,\mathbf{E}^\star(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating measured antenna electric fields over all possible pairs of $\mathbf{a}$ and $\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in matrix notation.

@@

We can then use the multiplication-convolution theorem to move the convolution in Equation~\ref{eqn:e-field-conv} to a square after the Fourier transform in equation~\ref{eqn:dirty-image-FX}.
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\mathbf{W}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna electric fields are weighted by the antenna noise, weighted and gridded by the antenna aperture kernel, Fourier transformed and finally squared to obtain the same image estimate that would have been obtained using equation~\ref{eqn:dirty-image-FX}.

@@

EPIC can accept dual-polarization inputs and produce images of all four instrumental cross-polarizations. Currently, two data input formats exist for reading in the electric field time samples measured by the antennas -- simulated electric fields based on a sky model using the simulator packaged with EPIC; and LWA data. Efforts to build interfaces for data from other telescopes are underway.

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The propagated electric fields are shown on the left at different time stamps, $t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by antennas are denoted by $\widetilde{E}_1(t)\ldots \widetilde{E}_\textrm{N}(t)$. The F-engine performs a temporal Fourier transform on the electric field time-series to obtain electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\mathbf{E}(\mathbf{a})$ in matrix notation) for each of the antennas. Each of the complex antenna gains are calibrated to correct the corresponding electric field spectra. These calibrated electric fields are gridded using an antenna-based gridding convolution function after which it is spatially Fourier transformed and squared to obtain an image cube for every time stamp. These images are then time-averaged to obtain the accumulated image $\mathcal{I}^\prime(f)$ ($\boldsymbol{\mathcal{I}^\prime}(\hat{\mathbf{s}})$ in matrix notation).

\begin{figure}
  \includegraphics[width=\columnwidth]{figure1}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{Flowchart of MOFF imaging in EPIC. The propagated electric fields shown on the left are measured as time-series $\widetilde{E}_1(t)\ldots \widetilde{E}_\textrm{N}(t)$ by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then Fourier transformed by the F-engine to produce electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series are imaged to produce images $\mathcal{I}^\prime_1(f)\ldots \mathcal{I}^\prime_\textrm{M}(f)$. These images are time-averaged to obtain the final image $\mathcal{I}^\prime(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure2}
  % \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{Flowchart of visibility-based software holographic imaging in EPIC. The FX process flow shares the F-engine with the MOFF process. Following the F-engine, the electric fields pass through the X-engine to produce visibilities $V_\textrm{ab}(f,t)$ which are calibrated and time-averaged. Then they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which are then Fourier transformed to obtain the image, $\mathcal{I}^\prime(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility-based imaging techniques. Time samples of electric fields measured by the antenna and digitized by the A/D converter is Fourier transformed to generate electric field spectra. This step can be parallelized across antennas. The output is then fed to either MOFF or visibility-based imaging pipelines.

@@

In a companion paper \citep{bea16}, we demonstrate a novel calibration technique (EPICal) which leverages the data products formed by direct imaging correlators to estimate antenna complex gains. This method correlates the antenna electric field signals with an image pixel from the output of the correlator in the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig.~\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration block). Furthermore it allows for arbitrarily complex sky models, and following the MOFF algorithm places no restriction on array layout, and accounts for non-identical antenna beam patterns. Direction dependent calibration can be achieved by correlating antenna signals with output pixels in the direction of $N_\textrm{c}$ calibration sources, then fitting for a functional model of the sky. Since antennas are only correlated with calibrator pixels, the computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$. The calibration performance of EPIC is explored for varying levels of completeness of sky model in \citet{bea16}.

@@

\par\medskip
\noindent {\bf Gridding Convolution}
\par\medskip
\noindent The antenna array aperture illumination over the entire grid, $W(\mathbf{r})$, is obtained by a projection of the individual antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  W(\mathbf{r}) &= \sum_a W_a(\mathbf{r}-\mathbf{r}_a),
\end{align}
or in matrix notation,
\begin{align}
  \mathbf{W}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathbf{I}(\mathbf{a}),
\end{align}
where, $\mathbf{I}(\mathbf{a})$ is a row of ones denoting coverage of the grid by kernel footprints of antennas. This is achieved by efficient multiplication with the sparse matrix created in the antenna-to-grid mapping process using the sparse matrices module in Python SciPy package. Unless $\mathbf{W}(\mathbf{r})$ includes time-dependent effects of the ionosphere or the instrument, it needs to be computed just once for the entire observation. However, the gridding of electric fields must be computed at every readout of the electric field spectra,
\begin{align}
  \mathbf{E}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathbf{E}(\mathbf{a}).
\end{align}

@@

In MOFF imaging, these are spatially Fourier transformed followed by a squaring operation at every time stamp for every frequency channel. In visibility-based imaging, the spatial Fourier transform is performed only once per integration time-scale and does not include a squaring operation.

\par\medskip
\noindent {\bf Time-averaging}
\par\medskip
\noindent In MOFF imaging, the measured antenna electric fields and the corresponding holographic electric field images are zero-mean stochastic quantities. Hence, they cannot be time-averaged to reduce noise. The statistical quantity stable with time in this case are the square of the holographic electric field images. Thus, squared images have to be formed at every instant of time before averaging as indicated in equation~\ref{eqn:dirty-image-MOFF}.

In contrast, visibilities measured by an antenna are statistically stable within an integration time interval. Hence, they are averaged after calibration as shown in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities before imaging because the repeated cost of spatial FFT can be avoided. Since this averaging has been performed already on the visibilities over an integration time-scale, the imaging step has to be performed only once per integration cycle.

@@

We use the EPIC simulator to generate stochastic electric field samples from a sky model consisting of 10 point sources of random flux densities $>10$~Jy each at random locations. In our simulations, we use 64 frequency channels each of width $\Delta f = 40$~kHz. The number of time stamps integrated in one integration cycle was kept at eight where each A/D timeseries is $1/\Delta f=25\,\mu$s long. We use the MWA array layout \citep{bea12} for demonstration. Only the inner 51 tiles within a square bounding box of 150~m on each side were used. We assumed all tiles are identical and have a square shaped electric field illumination footprint 4.4~m on each side. Besides the stochastic sky noise present in the simulated electric fields, no noise from the instrument is added.

@@

Before the outputs can be compared, we describe the elimination of a distinct difference between the two techniques. The squaring operation under MOFF imaging in the image plane introduces antenna auto-correlations around the zero spacing in the $uv$-plane which are absent in traditional visibility-based imaging. In order to facilitate a robust comparison between MOFF and visibility-based imaging techniques, these auto-correlations are removed from the MOFF algorithm output, which is otherwise not an essential part of the core algorithm. We describe below how they are removed.

@@

\subsection{Application to LWA data}\label{sec:LWA-data}

>>>>

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. \sout{It takes advantage of the multiplication-convolution theorem of Fourier transforms.} Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. EPIC is highly modular, parallelizable, implemented in object-oriented Python, and publicly available. We have verified the images produced to be equivalent to those from traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA1). {\bf We provide a detailed framework for imaging with heterogeneous arrays and show that EPIC} robustly estimates the input sky model for {\bf such} arrays. Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying an optimized version of EPIC. The algorithm is a strong candidate for instruments targeting transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena due to the availability of high-speed calibrated time-domain images and low output bandwidth relative to visibility-based systems.
\end{abstract}

@@

This paradigm shift requires a fundamentally new approach to the design of digital correlators \citep{lon00}. Modern correlators calculate the cross-power correlation between all antenna pairs in many narrow frequencies, forming \emph{visibilities}, the fundamental measurement of traditional radio interferometers. The computational requirements for a modern FX correlator scale with the number of antenna pairs, or the square of the number of antennas $\sim \Nant^2$ \citep{bun04}. For this reason traditional correlators have difficulty scaling to thousands of antennas. For example, the full HERA correlator for 352 dishes with 200 MHz of bandwidth requires 212 trillion complex multiply-accumulates per second (TMACS). Future arrays with thousands of collecting elements will require orders of magnitude more computation, making the correlator the dominant cost.

@@

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA1. {\bf In \S\ref{sec:versatility}, we present a mathematical framework for imaging with heterogeneous arrays and demonstrate the ability of EPIC} to robustly image data from such arrays. In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator is computationally more efficient, and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{i 2\pi f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \nonumber\\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-i 2\pi f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. The noise-like cross-terms ideally have zero mean. This expression can be further simplified using the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)$, as $\left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t = \mathcal{I}(\hat{\mathbf{s}},f,t)\,\delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{B}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power pattern, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier-transforming from the directional coordinates ($\hat{\mathbf{s}}$) to the measurement plane, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

@@

Hereafter, we adopt the matrix notation of \citet{mor11}, where vectors are represented with single coordinates, and matrices are represented by two coordinates denoting the spaces the operator transforms between. Since each frequency is processed independently, we drop the dependence on $f$. In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \mathbf{B}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}})$ is Fourier-transformed using $\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence function is weighted and summed using the antenna power response, $\mathbf{B}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \mathbf{E}(\mathbf{a})\,\mathbf{E}^\star(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating measured antenna electric fields over all possible pairs of $\mathbf{a}$ and $\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in matrix notation.

@@

We can then use the multiplication-convolution theorem to move the convolution in Equation~\ref{eqn:e-field-conv} to a \sout{square} {\bf product} after the Fourier transform in equation~\ref{eqn:dirty-image-FX}.
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\mathbf{W}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna electric fields are weighted by the antenna noise, weighted and gridded by the antenna aperture kernel, Fourier-transformed and finally squared to obtain the same image estimate that would have been obtained using equation~\ref{eqn:dirty-image-FX}.

@@

EPIC can accept dual-polarization inputs and produce images of all four instrumental cross-polarizations. Currently, two data input formats exist for reading in the electric field time samples measured by the antennas -- simulated electric fields based on a sky model using the simulator packaged with EPIC; and LWA1 data. Efforts to build interfaces for data from other telescopes are underway.

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The propagated electric fields are shown on the left at different time stamps, $t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by antennas are denoted by $\widetilde{E}_1(t)\ldots \widetilde{E}_\textrm{N}(t)$. The F-engine performs a temporal Fourier transform on the electric field time-series to obtain electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\mathbf{E}(\mathbf{a})$ in matrix notation) for each of the antennas. Each of the complex antenna gains are calibrated to correct the corresponding electric field spectra. These calibrated electric fields are gridded using an antenna-based gridding convolution function after which it is spatially Fourier-transformed and squared to obtain an image cube for every time stamp. These images are then time-averaged to obtain the accumulated image $\mathcal{I}^\prime(f)$ ($\boldsymbol{\mathcal{I}^\prime}(\hat{\mathbf{s}})$ in matrix notation).

\begin{figure}
  \includegraphics[width=\columnwidth]{figure1}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{Flowchart of MOFF imaging in EPIC. The propagated electric fields shown on the left are measured as time-series $\widetilde{E}_1(t)\ldots \widetilde{E}_\textrm{N}(t)$ by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then Fourier-transformed by the F-engine to produce electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series are imaged to produce images $\mathcal{I}^\prime_1(f)\ldots \mathcal{I}^\prime_\textrm{M}(f)$. These images are time-averaged to obtain the final image $\mathcal{I}^\prime(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure2}
  % \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{Flowchart of visibility-based software holographic imaging in EPIC. The FX process flow shares the F-engine with the MOFF process. Following the F-engine, the electric fields pass through the X-engine to produce visibilities $V_\textrm{ab}(f,t)$ which are calibrated and time-averaged. Then they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which are then Fourier-transformed to obtain the image, $\mathcal{I}^\prime(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility-based imaging techniques. Time samples of electric fields measured by the antenna and digitized by the A/D converter is Fourier-transformed to generate electric field spectra. This step can be parallelized across antennas. The output is then fed to either MOFF or visibility-based imaging pipelines.

@@

In a companion paper \citep{bea16}, we demonstrate a novel calibration technique (EPICal) which leverages the data products formed by direct imaging correlators to estimate antenna complex gains. This method correlates the antenna electric field signals with an image pixel from the output of the correlator in the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig.~\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration block). Furthermore, it allows for arbitrarily complex sky models, \sout{and following the MOFF algorithm} places no restrictions on array layout, and accounts for non-identical antenna beam patterns. Direction-dependent calibration can be achieved by correlating antenna signals with output pixels in the direction of $N_\textrm{c}$ calibration sources, then fitting for a functional model of the sky. Since antennas are only correlated with calibrator pixels, the computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$. The calibration performance of EPIC is explored for varying levels of completeness of sky model in \citet{bea16}.

@@

\par\medskip
\noindent {\bf Gridding Convolution}
\par\medskip
\noindent The antenna array aperture illumination over the entire grid, $W(\mathbf{r})$, is obtained by a projection of the individual antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  W(\mathbf{r}) &= \sum_a W_a(\mathbf{r}-\mathbf{r}_a),
\end{align}
or in matrix notation,
\begin{align}
  \mathbf{W}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathbf{I}(\mathbf{a}),
\end{align}
where, $\mathbf{I}(\mathbf{a})$ is a row of ones denoting coverage of the grid by kernel footprints of antennas. This is achieved by efficient multiplication with the sparse matrix created in the antenna-to-grid mapping process using the sparse matrix module in Python SciPy package. Unless $\mathbf{W}(\mathbf{r})$ includes time-dependent effects of the ionosphere or the instrument, it needs to be computed just once for the entire observation. However, the gridding of electric fields must be computed at every readout of the electric field spectra,
\begin{align}
  \mathbf{E}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathbf{E}(\mathbf{a}).
\end{align}

@@

In MOFF imaging, these are spatially Fourier-transformed followed by squaring at every time stamp for every frequency channel. In visibility-based imaging, the spatial Fourier transform is performed only once per integration time-scale and does not include a squaring operation.

\par\medskip
\noindent {\bf Time-averaging}
\par\medskip
\noindent In MOFF imaging, the measured antenna electric fields and the corresponding holographic electric field images are zero-mean stochastic quantities. Hence, they cannot be time-averaged to reduce noise. The statistical quantity stable with time in this case is the square of the holographic electric field image. Thus, squared images have to be formed at every instant of time before averaging as indicated in equation~\ref{eqn:dirty-image-MOFF}.

In contrast, visibilities measured by an antenna are statistically stable within an integration time interval. Hence, they are averaged after {\bf (and sometimes before)} calibration as shown in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities before imaging because the repeated cost of spatial FFT can be avoided. Since this averaging has been performed already on the visibilities over an integration time-scale, the imaging step has to be performed only once per integration cycle.

@@

We use the EPIC simulator to generate stochastic electric field samples from a sky model consisting of 10 point sources of random flux densities $>10$~Jy each at random locations. In our simulations, we use 64 frequency channels each of width $\Delta f = 40$~kHz. The number of time stamps integrated in one integration cycle was kept at eight where {\bf the duration of} each \sout{A/D} timeseries is $1/\Delta f=25\,\mu$s long. We use the MWA array layout \citep{bea12} for demonstration. Only the inner 51 tiles within a square bounding box of 150~m on each side were used. We assumed all tiles are identical and have a square-shaped electric field illumination footprint 4.4~m on each side. Besides the stochastic sky noise present in the simulated electric fields, no noise from the instrument is added.

@@

Before the outputs can be compared, we describe the elimination of a distinct difference between the two techniques. The squaring operation under MOFF imaging in the image plane introduces antenna auto-correlations around the zero-spacing in the $uv$-plane which are absent in traditional visibility-based imaging. In order to facilitate a robust comparison between MOFF and visibility-based imaging techniques, these auto-correlations are removed from the MOFF algorithm output, which is otherwise not an essential part of the core algorithm. We describe below how they are removed.

@@

\subsection{Application to LWA1 data}\label{sec:LWA-data}
